Quantifying Limits to Detection of Early Warning for Critical Transitions

### 2012-01-13 boettiger.tex
rewrites..

\subsection{The implicit model}
Foundational work on early warning signals has operated under the often-implicit assumption that the system dynamics contain a saddle-node bifurcation and looks for patterns that would be associated with approaching this bifurcation point.  Such methods

\subsection{Explicit use of the model}
Detection of warning signals as model choice (``non-parametric'' Carpenter \& Brock, unpublished Dakos, Kuehn arxiv, Dakos 2011?, Gross 2011)

>>>>

\subsection{The implicit model}
Foundational work on early warning signals has operated under the often-implicit assumption that the system dynamics contain a saddle-node bifurcation and looks for patterns that would be associated with approaching this bifurcation point.  The typical pattern has been a trend in a summary statistic such as variance, autocorrelation, skew, spectral ratio~\citep{Biggs2009}.  While attractive for their simplicity, such approaches are subject to numerous challenges.

1) While superficially simple, 
calculation of these statistics is subject to 
several somewhat arbitrary decisions and additional assumptions,
such as the choice of window size over which the statistic is computed,
the choice of detrending,
whether the consecutive averaging windows should overlap~\citep[\emph{e.g.}][]{Lenton2012}. % discusses these challenges, as does most of the others to lesser extent 
% ergodic problem

2) The warning signal, ``an increase in statistic $x$,'' is not quantitative,
making it difficult to attribute statistical significance to this statement.
While much of the literature employing these statistics simply avoids 
any attempt to quantify the significance of the detection, 
\citet{Dakos2008} relies on Kendall's correlation coefficient to quantify an increase,
and then goes through some pains to consider potential null models
that could be used to calculate statistical significance.  
Other measures of increase, such as Pearson's, have also been proposed.

3) Specifying an apprpriate null model for such a calculation is difficult.  
A supposed strength of the summary statistic approach 
the non-parametric significance test for Kendalls' tau tests against the null model that the sequence is independent,
something that is clearly violated when the statistic is computed in a sliding window.
Not surprisingly, \citet{Dakos2008} find this method overestimates the number of significant events relative to other nulls.  
Other non-parametric methods are tempting, 
such as shuffling the order of the data-points in the time-series,
but this breaks even the natural autocorrelation expected in a stable system that is not approaching any transition,
and thus should not be considered for a null.  
Relying on a more explicit but flexible model %or class of models
such as autogregressive processes is more consistent.  
% Does this go here?
We will see that such generic model-based approaches are becoming more common, 
and argue how they can be used to maximum advantage
% by simulation, likelihood fitting, and ROC curves.

% See Seekell et al for a summary stat with a good null

4) The underlying assumption that the system contains a saddle-node bifurcation
can be easilyoverlooked in summary-statistic based approach.  
For instance, variance may increase for a wide variety of reasons that 
do not signal an approaching transition~\citep{Schreiber2003, Schreiber2008} 
or may not increase as a bifurcation is approached~\citep{Livnia2012, Dakos2011a}.

% Does this section go here?
A system may experience a critical transition due to a rare fluctuation,
or a perturbation to the system state, %(Does this need a figure)?
rather than a bifurcation~\citep{Scheffer2001, Scheffer2009}.  
When applying detection approaches to data that have already experienced a critical transition,
we must condition on the 
and attempting to ascertain the cause of the transition
These are alternative null hypotheses 
 %(Dakos 2008, Lenton,  see Lenton 2009 transition, see 
\citep{Ditlevsen2010} % telling these two cases apart
\citep{Livina2011} noise induced transitions cannot be forecast.

5) Summary-statistic approaches inherently lack power. 
Methods for the detection of early warning signals are continually challenged by 
inadequate data  %\cite\cite\cite %often worse than we think -- as we will illustrate
After all, the challenge of extrapolating into the future across a dramatic shift is immense,
there is essentially no historical president for successfully \& reliably predicting regime shifts. % \cite Tetlock?
The signals one attempts to detect with these approaches are extremely weak.
Worse still, beyond this intuitive notion, there is little quantification of just how powerful our methods are,
even in artificial scenarios where uncertainy conforms to our parametric assumptions.  
% Seekell Am Nat, Contamin Eco Apps
We characterize the power

\subsection{Explicit use of the models}
Detection of warning signals as model choice (``non-parametric'' \citet{Carpenter2011e},  unpublished Dakos, Kuehn arxiv, Dakos 2011?, Gross 2011)
